Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-15 manuscript.tex
section 1 & 2

Theoretically, it was shown that a simple model may allow the propagation of such synfire chains~\citep{gewaltig_propagation_2001} by considering the dynamics of leaky integrate-and-fire (LIF) neurons in different groups of similar size. Each neuron of one group is connected by an excitatory synapse to the next. When a pulse is elicited in the first group, this may generate a spike in the next group. Depending on the concentration of synaptic weight values, this new activity may become more or less synchronized with respect to that of the previous pulse (as measured by the standard deviation of spike times within the pulse). Recursively applying this to a sequence of groups within a chain generates either a synfire propagation or not. A simple simulation of synfire propagation is shown in Figure~\ref{fig:diesman}. A crucial aspect of this emergence is explained by the dynamics of the spiking neuron model~\citep{gerstner_time_1995} and in particular the balance between excitation and inhibition~\citep{azouz_stimulus-selective_2008}. This balance was for instance modeled by feed-forward inhibition, a fine-scaled latency mechanism that is an essential ingredient in modelling so-called push-pull effects in the primary visual cortex~\citep{kremkow_push-pull_2016}. Further models have shown that such synfire chains could be embedded in topographies~\citep{aviel_embedding_2003}, while others used conductance-based neurons with feed-forward inhibition to improve the robustness of the propagation~\citep{kremkow_functional_2010}. In particular, this was implemented as a computational neuroscience benchmark model using the pyNN language~\citep{davison_pynn_2008} both in CPU-based and neuromorphic hardware~\citep{pfeil_six_2013}. % (see Figure~\ref{fig:pynn}).
%
%----------------------------%
%\begin{figure}
%\centering
%\includegraphics[width=.7\textwidth]{figures/fnins-07-00011-g004.jpg}
%\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background noise  is implemented as random Gaussian current. The state space gives %the duration of synfire chains as a function of the protocol's parameters.}\label{fig:pynn}
%\end{figure}
%----------------------------%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/Diesmann_et_al_1999.pdf}
\caption{\textbf{Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}.} The model consists of $10$ groups (arranged with the first group represented in the lowest row) of $100$ neurons each. Five pulses with decreasing jitter are generated in the first group around times $10$, $30$, $50$, $70$ and $90$~\ms~(with jitters given by a standard deviation which linearly decreases from $5$ to $1$~\ms). This generates a pulse after a certain processing delay in the second group with a different jitter. While the first two pulses progressively vanish, from the third input pulse, it is propagated to the following groups. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}
%----------------------------%

@@

In this section, we will review current evidence on how we may take advantage of spiking motifs in neurobiology, that is, in actual recordings from biological neural tissues. In most generic computational models, the neural activity is assumed to be encoded in the firing rate. For instance, the output of the so-called linear non-linear (L-NL) models is assumed to model the response of a biological neuron as the sequence of a linear integration followed by a non-linear spiking response generating spikes according to a Poisson point-process. As such, a simple decoding strategy is to infer the input knowing the neuron's tuning curves~\citep{jazayeri_optimal_2006} or simply by a simple regression~\citep{berens_fast_2012}. This latter model assumes a Bernoulli model for the generation of spikes, such that the decoding amounts to a single-layer logistic regression. An important perspective of these methods used to decipher the recorded activity is that it could be ultimately used to fit neural network models to the recorded activity~\citep{bellec_fitting_2021}. In this particular paper, authors fit summary statistics of neural data with a (differentiable) spiking network model. The loss function is the cross entropy (following Bernouilli hypothesis with a GLM where each unit is modelled with an SRM neuron~\citep{gerstner_time_1995}) and embedded with recurrent dynamics. In particular, it comes with \href{https://github.com/EPFL-LCN/pub-bellec-wang-2021-sample-and-measure}{code} and uses the publicly available \href{http://crcns.org/data-sets/vc/pvc-11}{V1-dataset} which allows supervising the model with the input being the movie and the output the spikes recorded. These type of model may infer sparse activity in a set of binary latent variables, each describing the activity of a cell assembly~\citep{warner_probabilistic_2022}.

>>>>

Theoretically, it was shown that a simple model may allow the propagation of such synfire chains~\citep{gewaltig_propagation_2001} by considering the dynamics of leaky integrate-and-fire (LIF) neurons in different groups of similar size. Each neuron of one group is connected by an excitatory synapse to the next. When a pulse is elicited in the first group, this may generate a spike in the next group. Depending on the concentration of synaptic weight values, this new activity may become more or less synchronized with respect to that of the previous pulse (as measured by the standard deviation of spike times within the pulse). Recursively applying this to a sequence of groups within a chain generates either a synfire propagation or not. A simple simulation of synfire propagation is shown in Figure~\ref{fig:diesman}. A crucial aspect of this emergence is explained by the dynamics of the spiking neuron model~\citep{gerstner_time_1995} and in particular the balance between excitation and inhibition~\citep{azouz_stimulus-selective_2008}. This balance was for instance modeled by feed-forward inhibition, a fine-scaled latency mechanism that is an essential ingredient in modelling so-called push-pull effects in the primary visual cortex~\citep{kremkow_push-pull_2016}. Further models have shown that such synfire chains could be embedded in topographies~\citep{aviel_embedding_2003}, while others used conductance-based neurons with feed-forward inhibition to improve the robustness of the propagation~\citep{kremkow_functional_2010}. In particular, this was implemented as a computational neuroscience benchmark model using the pyNN language~\citep{davison_pynn_2008} both in CPU-based and neuromorphic hardware~\citep{pfeil_six_2013}. % (see Figure~\ref{fig:pynn}).
%
%----------------------------%
%\begin{figure}
%\centering
%\includegraphics[width=.7\textwidth]{figures/fnins-07-00011-g004.jpg}
%\caption{A pyNN implementation of a Synfire chain with feedforward inhibition. The background noise  is implemented as random Gaussian current. The state space gives %the duration of synfire chains as a function of the protocol's parameters.}\label{fig:pynn}
%\end{figure}
%----------------------------%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/Diesmann_et_al_1999.pdf}
\caption{\textbf{Simulation of a synfire propagation \href{https://brian2.readthedocs.io/en/stable/examples/frompapers.Diesmann_et_al_1999.html}{using Brian}.} The model consists of $10$ groups (arranged with the first group represented in the lowest row) of $100$ neurons each. Five pulses with decreasing jitter are generated in the first group around times $10$, $30$, $50$, $70$ and $90$~\ms~(with jitters given by a standard deviation which linearly decreases from $5$ to $1$~\ms). This generates a pulse after a certain processing delay in the second group with a different jitter. While the first two pulses progressively vanish in the following groups, starting from the third input pulse (with a jitter of $3~\ms$), it is propagated to the following groups. This allows the propagation of the synchronous activity along the chain of the neural groups.}\label{fig:diesman}
\end{figure}
%----------------------------%

@@

In this section, we will review current evidence on how we may take advantage of spiking motifs in neurobiology, that is, in an effort to understand actual recordings from biological neural tissues. In most generic computational models, the neural activity is assumed to be encoded in the firing rate. For instance, the output of the so-called linear non-linear (L-NL) models is assumed to model the response of a biological neuron as the sequence of a linear integration followed by a non-linear spiking response generating spikes according to a Poisson point-process. As such, a simple decoding strategy is to infer the input knowing the neuron's tuning curves~\citep{jazayeri_optimal_2006} or simply by a simple regression~\citep{berens_fast_2012}. This latter model assumes a Bernoulli model for the generation of spikes, such that the decoding amounts to a single-layer logistic regression. An important perspective of these methods used to decipher the recorded activity is that it could be ultimately used to fit neural network models to the recorded activity~\citep{bellec_fitting_2021}. In this particular paper, authors fit summary statistics of neural data with a (differentiable) spiking network model. The loss function is the cross entropy (following Bernouilli hypothesis with a GLM where each unit is modelled with an SRM neuron~\citep{gerstner_time_1995}) and embedded with recurrent dynamics. In particular, it comes with \href{https://github.com/EPFL-LCN/pub-bellec-wang-2021-sample-and-measure}{code} and uses the publicly available \href{http://crcns.org/data-sets/vc/pvc-11}{V1-dataset} which allows supervising the model with the input being the movie and the output the spikes recorded. These type of model may infer sparse activity in a set of binary latent variables, each describing the activity of a cell assembly~\citep{warner_probabilistic_2022}.
